---
title: "Voter Registration"
author: "Joe Ciesielski"
date: "6/11/2018"
output: html_document
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(readxl)
library(here)
library(tidycensus)
library(stringr)
library(GGally)
library(stringr)
library(httr)
library(modelr)
library(forcats)
library(ggrepel)
library(rvest)
library(broom)

fix_names <- function(x) {
  names(x) <- tolower(make.names(names(x)))
  names(x) <- str_replace_all(names(x), "[/.]+", "_")
  x
}

make_names <- function(x) {
  x <- tolower(make.names(x))
  x <- str_replace_all(x, "[/.]+", "_")
  x
}

download_state_data <- function(x) {
  url <- paste0(
     "https://www2.census.gov/programs-surveys/demo/tables/voting/", 
     x, 
     ".xlsx"
   )
  GET(url, write_disk(tmp <- tempfile(fileext = ".xlsx")))
  df <- read_excel(tmp, skip = 1)
  names(df)[1] <- "variable"
  df %>% 
    fix_names() %>% 
    filter(!is.na(estimate)) %>% 
    slice(c(2:17, 19, 21, 23)) %>% 
    mutate(state = x)
}

capitalize <- function(x) {
  # capitalze all not conjunction words
  x <- tolower(x)
  x <- gsub("(?<!^)\\b(?:the|an?|[io]n|at|with(?:out)?|from|for|and|but|n?or|yet|[st]o|around|by|after|along|from|of)\\b(*SKIP)(*FAIL)|\\b(\\pL)", 
            "\\U\\1", x, perl = TRUE)
  x <- str_replace_all(x, " ", "")
  x
}

# state voter registration and voting info
state_voting <- "data/table04a.xls" %>% 
  here::here() %>% 
  read_excel(skip = 4) %>% 
  rename(state = X__1,
         total_pop = X__2,
         total_citizens = X__3) %>% 
  slice(2:52) %>% 
  mutate(state = capitalize(state)) %>% 
  fix_names()

# get partisanship of states
partisan <- "https://en.wikipedia.org/wiki/Political_party_strength_in_U.S._states" %>% 
  read_html() %>% 
  html_node(xpath = "/html/body/div[3]/div[3]/div[4]/div/table[2]") %>% 
  html_table() %>% 
  fix_names() %>% 
  rename(split = partisan_split_as_of_2014_update_)

partisan_split <- partisan %>% 
  select(state, split) %>% 
  mutate(
    split = str_extract_all(split, "\\d+\\.?\\d*"),
  )

partisan_split$split_low <- lapply(partisan_split$split, `[`, 2)
partisan_split$split_high <- lapply(partisan_split$split, `[`, 1)
partisan_split[c("split_low", "split_high")] <- lapply(partisan_split[c("split_low", "split_high")], as.numeric)

partisan_split$split_dif <- partisan_split$split_high - partisan_split$split_low

partisan_split <- partisan_split %>% 
  mutate(state = capitalize(state))

# get info on state voter id laws
id <- "https://en.wikipedia.org/wiki/Voter_ID_laws_in_the_United_States" %>% 
  read_html() %>%
  html_node(xpath = "/html/body/div[3]/div[3]/div[4]/div/table[2]") %>% 
  html_table() %>% 
  fix_names() %>% 
  mutate(
    state = capitalize(state),
    state = recode(state, "Washington,D.C." = "DistrictofColumbia"),
    # recode, mail only counts as no id
    id_law = case_when(
      state == "Iowa" ~ "No ID required", # Iowa enacted in 2017
      str_detect(type_of_law, "Strict") ~ "strict",
      str_detect(type_of_law, "non-strict") ~ "non-strict",
      TRUE ~ "No ID required"
    )
  )
# comine it all
drop_rows <- c("citizens|households")

state_demos <- state_voting$state %>%
  map_dfr(download_state_data) %>% 
  mutate(variable = make_names(variable)) %>% 
  dplyr::select(state, variable, percent_estimate) %>% 
  mutate(percent_estimate = as.numeric(percent_estimate)) %>% 
  spread(variable, percent_estimate) %>% 
  left_join(state_voting %>% select(state, 
                                    total_pop,
                                    percent_voted_citizen_,
                                    percent_voted_citizen_)) %>% 
  mutate_at(vars(percent_voted_citizen_, percent_voted_citizen_),
            function(x) x * .01) %>% 
  left_join(partisan_split %>% select(state, split_dif), by = "state") %>% 
  left_join(id %>% select(state, id_law), by = "state")

# na for split dif for DC; use mean
state_demos <- state_demos %>% 
  mutate(split_dif = if_else(is.na(split_dif), 
                             mean(split_dif, na.rm = TRUE), 
                             split_dif))

```

```{r}

state_demos %>% 
  ggplot(aes(x = fct_reorder(state, percent_voted_citizen_),
             y = percent_voted_citizen_)) +
    geom_point(stat = "identity") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



```{r}
state_demos %>% 
  select(bachelor_s_degree_or_higher,
         hispanic_or_latino,
         black_or_african_american_alone,
         asian_alone,
         households_with_income_100_000_or_more,
         x18_to_29_years,
         x65_years_and_over,
         below_poverty_level_poverty_rate_,
         split_dif,
         percent_voted_citizen_) %>% 
  ggpairs()
```

- outliers for Asian (Hawaii), Black (DC), and Latino (New Mexico)
- try interaction of ages, poverty level and income


```{r}

state_demos %>% 
  select(bachelor_s_degree_or_higher,
         hispanic_or_latino,
         black_or_african_american_alone,
         asian_alone,
         households_with_income_100_000_or_more,
         x18_to_29_years,
         x65_years_and_over,
         below_poverty_level_poverty_rate_,
         split_dif,
         percent_voted_citizen_) %>% 
  gather(predictor, value, 1:9) %>% 
  ggplot(aes(value, percent_voted_citizen_)) +
    geom_point() +
    facet_wrap(~predictor, scales = "free_x") +
    geom_smooth(method = "lm")
```

Impact of voter id laws
```{r}

ggplot(state_demos, aes(id_law, percent_voted_citizen_)) +
  geom_boxplot() +
  geom_jitter()

```

Voter id laws and minority

```{r}
ggplot(state_demos, aes(factor(id_law), white_alone)) + 
  geom_jitter() + 
  geom_smooth()

```


```{r}

ols <- lm(percent_voted_citizen_ ~ 
            hispanic_or_latino + 
            log(bachelor_s_degree_or_higher) +
            households_with_income_100_000_or_more  +
            log(split_dif) +
            id_law,
          data = state_demos)

summary(ols)
```

```{r}
resids <- state_demos %>% 
  add_predictions(ols) %>% 
  add_residuals(ols)
```

## Bootstrapped model
```{r}
bootstrapped <- state_demos %>% 
  select(percent_voted_citizen_,
         hispanic_or_latino,
         bachelor_s_degree_or_higher,
         households_with_income_100_000_or_more,
         split_dif,
         id_law) %>% 
  modelr::bootstrap(1000)

boot_model <- bootstrapped$strap %>% 
  map(~ lm(percent_voted_citizen_ ~ 
             hispanic_or_latino +
             log(bachelor_s_degree_or_higher) +
             households_with_income_100_000_or_more +
             log(split_dif) +
             id_law,
           data = .x)) %>% 
  map_df(broom::tidy, .id = ".id")

ggplot(boot_model, aes(estimate)) +
  geom_histogram() + 
  facet_wrap(~term, scales = "free_x")
```

```{r}
boot_model %>% 
  group_by(term) %>% 
  summarize(low = quantile(estimate, .025),
            high = quantile(estimate, .975)) %>% 
  knitr::kable()
```

## Deeper exploration of some states


```{r}
state_demos %>% 
  select(state,
         percent_voted_citizen_,
         hispanic_or_latino,
         bachelor_s_degree_or_higher,
         households_with_income_100_000_or_more,
         split_dif) %>% 
  gather(predictor, value, 3:6) %>% 
  ggplot(aes(fct_reorder(state, percent_voted_citizen_), value)) +
    geom_point() + 
    facet_wrap(~predictor, scales = "free_y") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
resids %>% 
  arrange(desc(abs(resid))) %>% 
  mutate(lab = if_else(row_number() %in% 1:5, 1, 0)) %>% 
  ggplot(aes(pred, percent_voted_citizen_)) +
    geom_point(aes(color = lab), show.legend = FALSE) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_text_repel(data = . %>% filter(lab == 1), 
                    aes(label = state))
```


 
Hawaii
- below expected
- recent reforms:
    - onine voter registration: http://editions.lib.umn.edu/electionacademy/2015/08/10/hawaii-launches-online-voter-registration/
    - same day registration: http://www.honolulumagazine.com/Honolulu-Magazine/Politics/June-2014/Same-Day-Voter-Registration-Coming-to-Hawaii/#.WyKzd6kna34
- historically low turnout
- photo id required
    
Mississippi
- last state in the nation to implement motor voter law, 
    - https://www.nytimes.com/1998/02/20/us/mississippi-legislature-accepts-bill-to-ease-voter-registration.html
    - what has changed? 
- politically engaged despite low bachelor degree attainment (lower than any state other than WV)
- strict voter id requirement implemented in 2014
  - https://ballotpedia.org/Voter_identification_laws_by_state 
- maybe Trump appeal? no, turnout down
  - https://mississippitoday.org/2016/11/21/election-recap-turnout-down-voters-were-prepared/ 
- 2016 first year without federal monitors
  - http://www.jacksonfreepress.com/news/2016/nov/22/mississippi-voter-turnout-down-over-last-president/ 

Wyoming
- similar turnout to 2012
  - http://soswy.state.wy.us/Elections/Docs/VoterProfile.pdf 
- allows same-day registration
- no id requirement

Tennessee
- id requirement
- historically low turnout
  - https://www.tennessean.com/story/news/2017/12/11/tennessee-voter-registration-cooper-dickerson-project-register/941325001/ 






