---
title: PA House - Progressive Above Replacement
author: Joseph Ciesielski
date: '2019-03-26'
slug: pa-house-progressive-above-replacement
categories: []
tags: []
comments: no
showcomments: yes
showpagemeta: yes
draft: true
---

```{r}

library(tidyverse)
library(broom)
library(jtcr)

```

State legislatures were not immune to the much discussed 'blue wave' of 2018. In Pennsylvania, a number of candidates ran on progressive and far left platfroms and won including Sara Innamoroto, Summer Lee, and Elizabeth Fiedler. 

A few months ago, Data for Progress put out a piece of US House Democrats fared on their '[Progressive above Replacement](https://www.dataforprogress.org/par-scores)' score. They predicted how progressive a Democrat should be based on characteristics of their district; progressive-ness was rated based on votes and co-sponsorships of bills in key progressive issues including public housing, tuition-free college, and renewable energy. 

Given the excitement about some of the far-left candidates for Pennsylvania's State House, I wanted to replicate this approach. 

I also wanted to see if scores on this metrics had any relationship to whether candidates were re-elected (or decided not to run) in 2018. 

## Methodology

I used votes on the following key divisive bills to asses progressiveness:

	- Bill 2154: rolling back regulations on oil and gas drilling
	- Bill 97: charter school reform
	- Bill 2060: taking away guns from domestic abusers
	- Bill 178: allowing teachers to be laid off based on evaluations
	- Bill 1419: expunging criminal records for those with minor offences
	- Bill 27: shielding names of officers involved in shootings
	- Bill 110: imposing limits on spending growth
	- Bill 2017: overturning Department of Revenue ruling disallowing deduction of business expenses
	- Bill 2477: funding medical marijuana research
	- Bill 2138: work requirements for Medicaid
	- Bill 2050: banning abortions for fetuses with Downs Syndrome 

We'll use four different data sets for this analysis. I scrpaed each from the web [add links] but saved each to my Github account. Let's read in each one at a time and take a look. 

```{r}

election_results <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/election_results.csv")

election_results

```
The first is the results from the 2016 PA House general election. I've already calculated the percent of votes that went for the Democrat. 

```{r}

bills <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/pa_house_bills_2017-18.csv")


bills
```

This data shows the bills, the topic, and whether or not a vote for the bill would be considered perspective. Next is how the reps votes on each of those bills ...

```{r}

votes <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/bill_votes.csv")

votes
```

... who won the 2018 election (so we can use this as an outcome) ...

```{r}

reps_2018 <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/reps_2018.csv")

reps_2018
```

... and a data set with the different name formats so we can combine the election results with the legislative data. 

```{r}

name_recode <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/votes_names.csv")

name_recode

```

Phew, ok. Now we can start merging some of the data together. 
