---
title: PA House - Progressive Above Replacement
author: Joseph Ciesielski
date: '2019-03-26'
slug: pa-house-progressive-above-replacement
categories: []
tags: []
comments: no
showcomments: yes
showpagemeta: yes
draft: true
---

```{r}

library(tidyverse)
library(broom)
library(jtcr)

win_share_voting <- read_csv("https://raw.githubusercontent.com/jtcies/pa-par/master/data/pahouse2016_votes_by_win_share.csv")

```

State legislatures were not immune to the much discussed 'blue wave' of 2018. In Pennsylvnaia, a number of candidates ran on progressive and far left platfroms and won including Sara Innamoroto, Summer Lee, and Elizabeth Fiedler. 

A few months ago, Data for Progress put out a piece of US House Democrats fared on their '[Progressive above Replacement](https://www.dataforprogress.org/par-scores)' score. They predicted how progressive a Democrat should be based on characteristics of their district; progressive-ness was rated based on votes and co-sponsorships of bills in key progressive issues including public housing, tuition-free college, and renewable energy. 

Given the excitement about some of the far-left candidates for Pennsylvania's State House, I wanted to replicate this approach. 

I also wanted to see if scores on this metrics had any relationship to whether candidates were re-elected (or decieded not to run) in 2018. 

## Methodology



```{r}




```



## The scores

## Impact on 2018 election