---
draft: true
---

```{r setup}

library(tidyverse)
library(nbastatR)

games <- game_logs(seasons = 2019)

game_1  <- games$idGame[[1]]

game_1_pbp <- play_by_play_v2(game_1)

game_1_pbp %>%
    View()

```
```{r g1firstaction}

game_1_first_action <- game_1_pbp %>%
    select(namePlayer1, namePlayer2, minuteGame, numberPeriod) %>%
    gather(drop, player, 1:2) %>%
    select(-drop) %>%
    group_by(player) %>%
    filter(minuteGame == min(minuteGame), !is.na(player)) %>%
    arrange(minuteGame) 

game_1_first_action %>%
    as.data.frame()
```
```{ri g1gt}

game_1_pbp %>% 
    fill(scoreAway, scoreHome, marginScore) %>%
    filter(minuteGame == 46.55) %>%
    select(scoreAway, scoreHome, marginScore)  %>%
    distinct()

```

```{r gtfun}

get_garbage_time_score <- function(game_pbp,
                                   # when to start looking for gt
                                   # default is start of the 4th quarter
                                   gt_start = 36,
                                   # number of players who first appearnce is after 
                                   # gt start
                                   n_players = 4
                                   ) {

    # excludes OT

    after_gt_start <- game_pbp %>%
        select(namePlayer1, namePlayer2, minuteGame) %>%
        gather(drop, player, 1:2) %>%
        select(-drop) %>%
        group_by(player) %>%
        filter(minuteGame == min(minuteGame), !is.na(player),
               minuteGame >= gt_start,
               minuteGame <= 48) %>%
        ungroup() %>%
        filter(row_number() >= n_players)

    if (nrow(after_gt_start) > 0) {

        gt <- unique(min(after_gt_start[["minuteGame"]]))

    } else {

        gt <- 48

    }
 
    gt_score <- game_pbp %>%
         fill(marginScore) %>%
         filter(minuteGame == gt) %>%
         distinct(idGame, marginScore, minuteGame) %>%
         mutate(n_players = n_players)
 
     gt_score

}

get_garbage_time_score(game_1_pbp)
```

```{r getgt}
set.seed(2019)

sample_games <- sample(games$idGame, 20)

games_pbp <- map(sample_games, play_by_play_v2)

garbage_time <- map_dfr(games_pbp, get_garbage_time_score)
```

```{r gtplot}

garbage_time %>%
    mutate(marginScore = abs(marginScore)) %>%
    ggplot(aes(minuteGame, marginScore)) +
        geom_point()
```

```{r cutoff}

n_players_max <- 6

n_players_list  <- rep(1:n_players_max, length(games_pbp))

games_rep_list <- rep(games_pbp, each = n_players_max)

mult_garbage_time <- map2_dfr(games_rep_list, n_players_list, ~ get_garbage_time_score(.x, n_players = .y))

mult_garbage_time %>%
    ggplot(aes(minuteGame, abs(marginScore), color = factor(n_players))) +
        geom_point() +
        facet_wrap(~idGame)

```

It looks like keeping it at least two players whose first action is in the fourth quarter gets us a pretty good defintion of garbage time. 



## Other questions?

- What are other ways to define garbage time start (than a player's first aciton)? 
- How do coaches differ in whether or not they go into garbage time mode? 
- Has the boundry for garbage time changed over time? 
- How close does the score have to get for a coach to put their starters back in?
- How to factor in games going to overtime? 

