---
draft: true
---

```{r setup}

library(tidyverse)
library(nbastatR)
library(rethinking)

games <- game_logs(seasons = 2019) %>%
    distinct(idGame)

game_1  <- games$idGame[[1]]

game_1_pbp <- play_by_play_v2(game_1)

game_1_pbp %>%
    View()

```
```{r g1firstaction}

game_1_first_action <- game_1_pbp %>%
    select(namePlayer1, namePlayer2, minuteGame, numberPeriod) %>%
    gather(drop, player, 1:2) %>%
    select(-drop) %>%
    group_by(player) %>%
    filter(minuteGame == min(minuteGame), !is.na(player)) %>%
    arrange(minuteGame) 

game_1_first_action %>%
    as.data.frame()
```
```{ri g1gt}

game_1_pbp %>% 
    fill(scoreAway, scoreHome, marginScore) %>%
    filter(minuteGame == 46.55) %>%
    select(scoreAway, scoreHome, marginScore)  %>%
    distinct()

```

```{r gtfun}

get_garbage_time_score <- function(game_pbp,
                                   # when to start looking for gt
                                   # default is start of the 4th quarter
                                   gt_start = 36,
                                   # number of players who first appearnce is after 
                                   # gt start
                                   n_players = 2
                                   ) {

    # excludes OT

    after_gt_start <- game_pbp %>%
        select(namePlayer1, namePlayer2, minuteGame) %>%
        gather(drop, player, 1:2) %>%
        select(-drop) %>%
        group_by(player) %>%
        filter(minuteGame == min(minuteGame), !is.na(player),
               minuteGame >= gt_start,
               minuteGame <= 48) %>%
        ungroup() %>%
        filter(row_number() >= n_players)

    if (nrow(after_gt_start) > 0) {

        gt <- unique(min(after_gt_start[["minuteGame"]]))

    } else {

        gt <- 48

    }
 
    gt_score <- game_pbp %>%
         fill(marginScore) %>%
         filter(minuteGame == gt) %>%
         arrange(desc(abs(marginScore))) %>%
         distinct(idGame, minuteGame, .keep_all = TRUE) %>%
         select(idGame, minuteGame, marginScore) %>%
         mutate(
            n_players = n_players,
            garbage = if_else(minuteGame < 48, 1L, 0L)
        )
 
     gt_score

}

get_garbage_time_score(game_1_pbp)
```

```{r getgtsample}
set.seed(2019)

sample_games <- sample(games$idGame, 20)

sample_games_pbp <- purrr::map(sample_games, play_by_play_v2)

n_players_max <- 6

n_players_list  <- rep(1:n_players_max, length(sample_games_pbp))

games_rep_list <- rep(sample_games_pbp, each = n_players_max)

mult_garbage_time <- map2_dfr(
    games_rep_list, n_players_list, 
    ~ get_garbage_time_score(.x, n_players = .y)
)

mult_garbage_time %>%
    filter(garbage == 1) %>%
    ggplot(aes(minuteGame, abs(marginScore), 
               group = idGame,
               color = factor(n_players))) +
        geom_point(size = 2) +
        geom_line(show.legend = FALSE)

```

Occasionally a team will bring in a player for their first action in the first quarter for tactical reasons. But it looks like by the time a second player sees their first action in the fourth quarter, we can consider garbage time to have offically started. The margin can increase after garbage time starts, so we don't want to overestimate the point at which garbage time starts. 

```{r getgt}

game_ids <- games$idGame

games_pbp <- purrr::map(game_ids, play_by_play_v2)

games_pbp <- bind_rows(games_pbp)

write_csv(games_pbp, here::here("content/data/nba_pbp_2019.csv"))

```




## Other questions?

- What are other ways to define garbage time start (than a player's first aciton)? 
- How do coaches differ in whether or not they go into garbage time mode? 
- Has the boundry for garbage time changed over time? 
- How close does the score have to get for a coach to put their starters back in?
- How to factor in games going to overtime? 

