---
title: PA House - Progressive Above Replacement
author: Joseph Ciesielski
date: '2019-10-21'
slug: pa-house-progressive-above-replacement
categories: []
tags: []
comments: no
showcomments: yes
showpagemeta: yes
draft: true
---

State legislatures were not immune to the much discussed 'blue wave' of 2018. In Pennsylvania, a number of candidates ran on progressive and far left platfroms and won including Sara Innamoroto, Summer Lee, and Elizabeth Fiedler. 

Earlier this year, Data for Progress put out a piece of US House Democrats fared on their '[Progressive above Replacement](https://www.dataforprogress.org/par-scores)' score. They predicted how progressive a Democrat should be based on characteristics of their district; progressive-ness was rated based on votes and co-sponsorships of bills in key progressive issues including public housing, tuition-free college, and renewable energy. 

I set out to replicate this analysis. In addition, I wanted to see if scores on this metrics had any relationship to whether candidates were re-elected (or decided not to run) in 2018. 

```{r}

library(tidyverse)
library(rvest)
library(broom)
library(scales)
library(rethinking)
library(plotly)

theme_set(theme_minimal())

```

## Methodology

I used votes on the following key divisive bills to asses progressiveness:

	- Bill 2154: rolling back regulations on oil and gas drilling
	- Bill 97: charter school reform
	- Bill 2060: taking away guns from domestic abusers
	- Bill 178: allowing teachers to be laid off based on evaluations
	- Bill 1419: expunging criminal records for those with minor offences
	- Bill 27: shielding names of officers involved in shootings
	- Bill 110: imposing limits on spending growth
	- Bill 2017: overturning Department of Revenue ruling disallowing deduction of business expenses
	- Bill 2477: funding medical marijuana research
	- Bill 2138: work requirements for Medicaid
	- Bill 2050: banning abortions for fetuses with Downs Syndrome 

### Data wrangling

First we'll scrape results from the 2016 and 2018 elections. 

```{r}

scrape_election <- function(year) {
  
 sprintf(
   "https://en.wikipedia.org/wiki/%i_Pennsylvania_House_of_Representatives_election",
   year
 ) %>% 
    read_html() %>%
    html_table(fill = TRUE) %>%
    .[[19]] %>%
    janitor::clean_names() %>%
    tbl_df() %>%
    select(-party, -party_2, -party_3, -percent) %>%
    rename(party = party_4) %>% 
    mutate(
      year = year,
      votes = parse_number(votes)
    )

}

election_results <- c(2016,  2018) %>% 
  map_dfr(scrape_election) %>% 
  group_by(district, year) %>% 
  mutate(percent = votes / sum(votes)) %>% 
  ungroup()

```

Next we'll get the information on the bills ...

```{r}

bills <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/pa_house_bills_2017-18.csv")


bills
```

... and votes on those bills[^1].

```{r}

votes <- read_csv("https://raw.githubusercontent.com/jtcies/jtcies_site/master/content/data/pa-par/bill_votes.csv")

votes
```

Phew, ok. Now we can start merging some of the data together. The goal is create a data set that inlcude election results, voting record, and whether someone was still a rep in 2018. 

First pull out the election winners. No independents won in 2016, so that makes this a little easier.

```{r}

election_tidy <- election_results %>%
    group_by(district, year) %>%
    mutate(
		win = percent == max(percent),
		margin = (percent - lead(percent, default = 0)) * 100
	) %>%
	filter(win) %>% 
  ungroup() %>% 
  rename(representative = candidate)

election_both <- election_tidy %>% 
  filter(year == 2016) %>% 
  left_join(
    election_tidy %>% 
      filter(year == 2018), 
    by = "district",
    suffix = c("_2016", "_2018")
    ) %>% 
  mutate(same = representative_2016 == representative_2018)

election_both
```

Next we'll create a summary of each representative's voting behavior by 

```{r}

prog_summary <- votes %>% 
    left_join(bills, by = "bill") %>%
    mutate(
      progressive_vote = case_when(
        progressive == 1 & vote == "yay" ~ 1,
        progressive == -1 & vote == "yay" ~ 0,
        progressive == 1 & vote == "nay" ~ 0,
        progressive == -1 & vote == "nay" ~ 1
      )
    ) %>% 
  group_by(rep) %>% 
  summarise(
    progressive_votes = sum(progressive_vote),
	progressive_avg = mean(progressive_vote),
	votes_cast = n()
  )

```

Merge the data.

```{r}

election_merge <- election_both %>%
	separate(representative_2016, into = c("first", "last", "suffix"), sep = " |-",
	         remove = FALSE) %>%
	mutate_at(vars(first, last, suffix), tolower) %>%
	group_by(last) %>%
	mutate(
		merge_name = case_when(
			!is.na(suffix) ~ suffix,
			length(row_number()) > 1 ~ paste0(last, " ", str_sub(first, 1, 1)),
			TRUE ~ last
		),
		merge_name = case_when(
			merge_name == "keller m" ~ "keller mk",
			merge_name == "keller b" ~ "keller w",
			merge_name == "braneky" ~ "krueger",
			merge_name == "quinn" & str_detect(last, "marguerite") ~ "quinn m",
			merge_name == "quinn" ~ "quinn c",
			merge_name == "day g" ~ "day",
			TRUE ~ merge_name
		)
	) %>%
	ungroup()


prog_merge <- prog_summary %>%
	mutate(rep = tolower(str_replace_all(rep, ",|\\.", "")))


pa_house_tidy <- election_merge %>%
	left_join(prog_merge, by = c("merge_name" = "rep"))

```

## Visualizing

First let's look at margin of victory by party. We should have 121 Repbublicans and 82 Democrats. 

```{r}

pa_house_tidy %>% 
  count(party_2016)

```

```{r}
pa_house_tidy %>% 
  ggplot(aes(margin_2016, fill = party_2016)) +
    geom_histogram(binwidth = 2, show.legend = FALSE) +
    facet_wrap(~party_2016) +
    scale_fill_manual(
      values = c("Democratic" = "blue",
                 "Republican" = "red")
    ) +
    labs(
			x = "margin of victory - 2016 election"
    )
    

```

It looks like both parties had an equal number of wins in uncontested districts, but Democrats had a greater percentage (because they held fewer seats).

```{r}
pa_house_tidy %>%
  group_by(party_2016) %>% 
  summarise(
    `count uncontested` = sum(margin_2016 == 100),
    `percent uncontested` = percent(mean(margin_2016 == 100))
  ) %>% 
  knitr::kable()
 
```

```{r}

pa_house_tidy %>% 
  group_by(party_2016) %>% 
  summarise(
    `count incumbent win` = sum(representative_2016 == representative_2018),
    `count flipped party` = sum(party_2016 == party_2018)
  ) %>% 
  knitr::kable()

```

```{r}

pa_house_tidy %>% 
  ggplot(aes(margin_2016, margin_2018, color = party_2018)) +
    geom_jitter(size = 3) +
    facet_wrap(~ party_2016, ncol = 1) +
     scale_color_manual(
      values = c("Democratic" = "blue",
                 "Republican" = "red")
    ) +
    labs(
        color = "Party of district\nafter 2018 election",
        x = "margin of victory - 2016 election",
        y = "margin of victory - 2018 election"
    )


```
Here we can see some of the results of the 'blue wave'. Many districts that went solidly Republican in 2016 were closer in 2018 - those are seen along the right hand sand of the bottom half of the chart. Similary, many closely contest districts went entirely for Democrats in 2018 - those are along the top of the upper chart. You can alos see a clump of blue dots among the bottom half representing those districts that flipped from Republican to Democrat.


### Progressiveness by win margin

```{r mod}
pa_house_tidy <- pa_house_tidy %>%
    filter(!is.na(progressive_avg))

model_dat <- list(
    votes = pa_house_tidy$votes_cast,
    progressive_votes = as.integer(pa_house_tidy$progressive_votes),
    party = as.integer(as.factor(pa_house_tidy$party_2016)),
    margin = scale(pa_house_tidy$margin_2016)[,1]
)

glmod <- ulam(
    alist(
        progressive_votes ~ dbinom(votes, p),
        logit(p) <- a[party] + b[party] * margin,
        a[party] ~ dnorm(a_bar, a_sigma),
        a_bar ~ dnorm(0, 1.5),
        a_sigma ~ dexp(1),
        b[party] ~ dnorm(0, b_sigma),
        b_sigma ~ dexp(1)
    ), data = model_dat,
    chains = 4, cores = 4, log_lik = TRUE
)


precis(glmod, depth = 2)

pairs(glmod)
```

```{r modpredplot}

post <- extract.samples(glmod)

apply_preds <- function(margin, party) {

    post$a[, party] + post$b * margin


} 

possible <- crossing(margin = seq(-1, 1, length.out = 100),
         party = c(1, 2))

preds <- purrr::map2(possible$margin, possible$party, apply_preds)
```


I'll come back to the shaded part in a second. 

There is definitely a trend that shows Democrats as more progressive and Republicans as less progressive if they won their district by a greater margin. However, it's not a huge magnitude. A Democrat who won by less than one point is expeted to have about a .73 progressive average, compared to a Democrat winning uncontested, who is expected to vote progressive about 82% of the time. 

## Whos is most and least progressive based on their margin of victory? 

The original PAR article looked at Democrats who were less progressive than expected based on their margin of victory. So who are they in PA? 


```{r}
dems_less_progressive <- augmented %>% 
  filter(
    party_2016 == "Democratic", margin_2016 == 100,
    progressive_avg < .fitted
  )
```

There are `r nrow(dems_less_progressive)` Democrats who won uncontested in 2016 races who were less progressive than expected. Let's look at some of them. 

```{r}
dems_less_progressive %>% 
  arrange(progressive_avg) %>% 
  slice(1:10) %>% 
  select(district, party_2016, representative_2016, margin_2016,
         progressive_avg, party_2018, representative_2018, margin_2018)

```

All were re-elected, and all but one had uncontested general elections in 2018 (although some may have been primaried). 

Chris Sainato has is the represenative for New Castle in Western PA and has been in office since 1995. However, he received a challenge from Republican in 2018 and won by only 13 points. So perhaps he knew that this was coming. He also released an (op-ed)[https://www.ncnewsonline.com/opinion/columns/chris-sainato-restoring-civility-to-our-politics/article_9e82083d-e78d-58e2-8527-4957bc9710b0.html] about civility in politics, so he's definitely going all-in on the moderate Dem approach. 

Tony DeLuca is has been in office since 1983 has been the subject of multiple investiagtions related to issues related to nepotism and fraud, was the subject of a recent BuzzFeed article about his relationship with insurance compaines. Safe to say he's not the most progressive guy in the House. 

## What about the Republicans?

Let's come back to that grey shaded area in the chart. This shows Republicans who had a margin of victory below 31 points. It looks like very few dots above the line are outlined in black - meaning those candidates were not re-elected - compared to below the line. The table below shows the breakdown of what happened to these representatives in the 2018 election. 

```{r}
 augmented %>% 
     filter(party_2016 == "Republican", margin_2016 < 50) %>%
     group_by(negative = if_else(.fitted < progressive_avg, 
                                 "more progressive than expected", 
                                 "less progressive than expected")) %>%
     count(status_2018) %>%
     spread(negative, n, fill = 0) %>%
     rename(`2018 result` = status_2018)
```

Of the 24 Republicans who were more progressive than the model predicts, 8 were defeated and 8 retired. Of the 45 less progressive than predicted, just two retired and none were defeated. 

This is farily staggering. Even in a year when things swung toward more liberal candidates, Republcians that showed some amount of moderation were heavily punished for it. 

Perhaps this is another example of increased partisanship? I'm not sure. 

## Conclusions

I set out to replicate the methodology used by Data for Progress. The goal of that analysis is to identify Democrats who could be pushed to support more progressive issues because they have little chance of being outsted. We that this is true for Pennsylvania as well, but we also saw examples of seats that seemed same become contested. 

However, I found the ways in which Repbulicans who showed even a little bit of progressiveism were punished to be the most interesting finding of this analysis. Perhaps this is just a symtpom of the 'blue wave' - those seats were vulnerable and Democrats pounced. It seems like it can't be just this though; when you look similarly contested seats, Republicans who were the least progressive were re-elected while those who were more progressive were not re-elected. 

[^1]: If you want to see the code I used to scrape this data, it's [here](https://www.github.com/jtcies/pa-par)

